An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-08-18 latex_files/ch2/2_2_LFI_background.tex
Michael's edits on overleaf, LFI background and OMC

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample, or simulate, data points, but that we cannot
evaluate the likelihood of a specific set of observations
$\data$. Formally, a simulator-based model is a parameterised family
of probability density functions $\{ p(\yb|\thetab) \}_{\thetab}$,
whose closed-form is either unknown or computationally intractable to
evaluate. Whereas evaluating $p(\yb|\thetab)$ is intractable, sampling
is feasible. Practically, a simulator can be understood as a black-box
data generating machine $M_r$\footnote{The subscript $r$ in $M_r$
indicates the \textit{random} simulator. In the next chapters we will
introduce $M_d$ which stands for the \textit{deterministic}
simulator.} that generates samples $\yb$ in a stochastic manner for
any given a set of parameters $\thetab$.% i.e.\
$M_r(\thetab) \rightarrow \yb$.

@@

The approximation in \eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractabilities of the likelihood
function in \eqref{eq:likelihood}. Other strategies include modelling
the (stochastic) relationship between $\thetab$ and $\yb$, and its
reverse, or framing likelihood-free inference as a ratio estimation
problem \citep[see e.g.\ ][]{Blum2010, Wood2010, Papamakarios2016,
Thomas2016, Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both OMC and robust OMC that we introduce next
are based on the approximation in \eqref{eq:approx_likelihood}.

>>>>

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample, or simulate, data points, but that we cannot
evaluate the likelihood of a specific set of observations
$\data$. Formally, a simulator-based model is a parameterised family
of probability density functions $\{ p(\yb|\thetab) \}_{\thetab}$,
whose closed-form is either unknown or computationally intractable to
evaluate. Whereas evaluating $p(\yb|\thetab)$ is intractable, sampling
is feasible. Practically, a simulator can be understood as a random black-box
machine (computer code) $M_r$ that generates samples $\yb$ in a stochastic manner for
any given a set of parameters $\thetab$. We can isolate the randomness by introducing the stochastic nuisance variables $\vb \sim p(\vb)$ so that the simulator becomes a deterministic mapping $M_d$ that maps $(\thetab, \vb)$ to the data $\yb$. The distribution $p(\vb)$ is determined by the random draws performed when running $M_r$. Importantly, we can fix the value of $\vb$ while changing $\thetab$ by fixing the random seed used with $M_r$.

@@

The approximation in \eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractabilities of the likelihood
function in \eqref{eq:likelihood}. Other strategies include modelling
the (stochastic) relationship between $\thetab$ and $\yb$, and its
reverse, or framing likelihood-free inference as a ratio estimation
problem \citep[see e.g.\ ][]{Blum2010, Wood2010, Papamakarios2016,
Thomas2016, Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both OMC and robust OMC, which we introduce next,
are based on the approximation in \eqref{eq:approx_likelihood}.

### 2021-08-18 latex_files/ch2/2_3_OMC.tex
Michael's edits on overleaf, LFI background and OMC

For describing the likelihood approximation introduced in OMC, it is
convenient to firstly define the indicator function
$\indicator{\region(\yb)}(\xb)$, that returns 1 if
$\xb \in \region(\yb)$ and 0 otherwise. If $d(\cdot,\cdot)$ is a
formal distance, due to symmetry
$\indicator{\region(\yb)}(\xb) = \indicator{\region(\xb)}(\yb)$, so
the expressions can be used interchangeably.

\begin{gather} \label{eq:indicator} \indicator{\region(\yb)}(\xb)=
  \left\{
    \begin{array}{ll}
      1 & \mbox{if } \xb \in \region(\yb) \\
      0 & \mbox{otherwise} 
    \end{array} \right. \end{gather}

\noindent
Based on equation~\eqref{eq:approx_likelihood} and the indicator
function~\eqref{eq:indicator}, we can approximate the likelihood as:

\begin{gather} \label{eq:approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) =
  \int_{\yb \in B_\epsilon(\data)}p(\yb|\thetab)d\yb =
  \int_{\yb \in \R^D} \indicator{\region(\data)}(\yb)p(\yb|\thetab)d\yb\\
  \approx \frac{1}{N} \sum_i^N \indicator{\region(\data)}(\yb_i),\text{ where }
  \yb_i \sim M_r(\thetab) \label{eq:init_view}\\
  \approx \frac{1}{N} \sum_i^N \indicator{\region (\data)} (\yb_i)
  \text{ where } \yb_i = M_d(\thetab, \vb_i), \vb_i \sim p(\vb) \label{eq:alt_view}
\end{gather}
%
The procedure is quite intuitive; approximating the likelihood of a
specific $\thetab$ requires sampling from the data generator and count
the fraction of samples that lie inside the area around the
observations. Nevertheless, by using the approximation of equation
\eqref{eq:init_view} we need to draw $N$ new samples for each distinct
evaluation of $L_{d,\epsilon}(\thetab)$; this makes this approach
quite inconvenient from a computational point-of-view. For this
reason, OMC approximates the integral as in equation
\eqref{eq:alt_view}; the nuisance variables, $\vb_i \sim p(\vb)$, are
sampled once and we count the fraction of samples that lie inside the
area using the deterministic simulators
$M_d(\thetab, \vb_i) \: \forall i$. Hence, the evaluation
$L_{d,\epsilon}(\thetab)$ for each different $\thetab$ does not imply
drawing new samples all over again. Based on this approach, the
unnormalised approximate posterior can be defined as:

\begin{equation} \label{eq:approx_posterior_omc}
  p_{d,\epsilon}(\thetab|\data)
  \propto p(\thetab) \sum_i^N \indicator{ \region(\data)} (\yb_i)
\end{equation}

>>>>

Our description of OMC follows \citet{Ikonomov2019} who base their explanation of OMC on the approximate likelihood function in \eqref{eq:approx_likelihood}. With the indicator function
%
\begin{equation} \label{eq:indicator} \indicator{\region(\yb)}(\xb)=
  \left\{
    \begin{array}{ll}
      1 & \mbox{if } \xb \in \region(\yb) \\
      0 & \mbox{otherwise} 
    \end{array} \right. \end{equation}
%
we can write the approximate likelihood function $L_{d, \epsilon}(\thetab)$ in \eqref{eq:approx_likelihood} as
\begin{gather} \label{eq:approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data)) =
  \int_{\yb \in B_\epsilon(\data)}p(\yb|\thetab)d\yb =
  \int_{\yb \in \R^D} \indicator{\region(\data)}(\yb)p(\yb|\thetab)d\yb
\end{gather}
which we can approximate as a sample average
\begin{equation}
L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\region (\data)} (\yb_i)
  \text{ where } \yb_i = M_d(\thetab, \vb_i), \quad \vb_i \sim p(\vb). \label{eq:alt_view}
\end{equation}
We thus approximate the likelihood of a
specific $\thetab$ by counting the fraction of samples that lie inside the area around the
observations. The crucial point in \eqref{eq:alt_view} is that we may only sample the nuisance variables $\vb_i \sim p(\vb)$ once and use the sample set of samples to approximate $L_{d,\epsilon}(\thetab)$ for different values of $\thetab$. Based on this approach, we introduce the approximate posterior
\begin{equation} \label{eq:approx_posterior_omc}
  p_{d,\epsilon}(\thetab|\data)
  \propto p(\thetab) \sum_i^N \indicator{ \region(\data)} (\yb_i) \text{ where } \yb_i = M_d(\thetab, \vb_i), \quad \vb_i \sim p(\vb).
\end{equation}
